---
title: "Tutorial: rbind"
date: "10/04/2020"
output: html_document
---

<!--
```{r cache=FALSE, include=FALSE}
knitr::read_chunk('rbind_tutorial.R')
```
-->
```{r setup, include=FALSE}
colorize <- function(x, color) {
  if (knitr::is_latex_output()) {
    sprintf("\\textcolor{%s}{%s}", color, x)
  } else if (knitr::is_html_output()) {
    sprintf("<span style='color: %s;'>%s</span>", color, 
      x)
  } else x
}
```

```{r include=FALSE}
source("rbind_tutorial.R", echo = FALSE, keep.source = TRUE)
```

# rbind in R | 3 Examples (Vector, Data Frame & rbind.fill for Missing Columns)


**Basic R Syntax:**
```{r eval=FALSE}
rbind(my_data, new_row)
```

The name of the _rbind_ R function stands for _row-bind_. The rbind function can be used to combine several vectors, matrices and/or data frames by rows. Above, you can find the basic code for rbind in R.

In the following article, I’m going to provide you with 3 examples for the application of the rbind function in R. Let’s start right away.

## Example 1: rbind Vector to Data Frame

The easiest way of using rbind [in R](https://statisticsglobe.com/r-programming-language/) is the combination of a vector and a data frame. First, let’s create some example data frame…

```{r eval=FALSE}
x1 <- c(7, 4, 4, 9)                  # Column 1 of data frame
x2 <- c(5, 2, 8, 9)                  # Column 2 of data frame
x3 <- c(1, 2, 3, 4)                  # Column 3 of data frame
data_1 <- data.frame(x1, x2, x3)     # Create example data frame
```

…and an example vector:

```{r eval=FALSE}
vector_1 <- c(9, 8, 7)               # Create example vector
```

Now, let’s rbind this vector to the data frame:

```{r eval=TRUE, echo=TRUE}
rbind(data_1, vector_1)              # rbind vector to data frame
```

The above table illustrates the output of the rbind function: 
* The first four rows are identical to our original data frame data_1; 
* the fifth row is identical to our vector vector_1.
 
## Example 2: rbind Two Data Frames in R

The rbind command can also be applied to two data frames. Let’s create a second data frame and row bind it to data_1 (the data frame that we created above):

```{r eval=FALSE}
x1 <- c(7, 1)                        # Column 1 of data frame 2
x2 <- c(4, 1)                        # Column 2 of data frame 2
x3 <- c(4, 3)                        # Column 3 of data frame 2
data_2 <- data.frame(x1, x2, x3)     # Create second data frame
```

We can rbind these two data frames with the same R code as before:

```{r eval=TRUE, echo=TRUE}
rbind(data_1, data_2)                # rbind two data frames in R
```

As in Example 1, the upper part of the rbind output consists of data_1 and the lower part of the rbind output consists of data_2.

**Note:** Both data frames have the same column names. When the columns of the two data frames differ, it gets a bit more complicated. In the next example, I’m going to show you how to rbind data frames with different column names.

## Example 3: rbind fill – Row Bind with Missing Columns

The binding of data frames with different columns / column names is a bit more complicated with the rbind function. R usually returns the error “Error in match.names(clabs, names(xi))”, if you try to use the rbind function for data frames with different columns.

For that reason, the [plyr package](https://cran.r-project.org/web/packages/plyr/plyr.pdf) (be careful: it’s called plyr; not dplyr) provides the rbind.fill R function as add-on to the rbind base function. In the following example, I’ll show you how to use the plyr rbind.fill [function in R](https://statisticsglobe.com/r-functions-list/).

Consider the following two example data frames:

```{r eval=TRUE, echo=TRUE}
col1 <- c(5, 1, 1, 8)                # Column 1 of data frame 1 plyr example
col2 <- c(9, 7, 5, 1)                # Column 2 of data frame 1 plyr example
col3 <- c(1, 1, 2, 2)                # Column 3 of data frame 1 plyr example
 
data_plyr1 <- data.frame(col1, col2, col3) # Create plyr data frame 1
```
```{r eval=TRUE, echo=TRUE}
col3 <- c(5, 1, 1, 8)                # Column 1 of data frame 2 plyr example
col4 <- c(9, 7, 5, 1)                # Column 2 of data frame 2 plyr example
 
data_plyr2 <- data.frame(col3, col4) # Create plyr data frame 2
```

To apply rbind.fill in R, we need to load the [plyr package](https://statisticsglobe.com/plyr-r-package) first:

```{r}
library("plyr") 
```

Now, we can apply the rbind.fill R command to our two example data frames:

```{r eval=TRUE, echo=TRUE}
rbind.fill(data_plyr1, data_plyr2)   # Apply the rbind.fill plyr function
```

The above table makes it clear how rbind fill works: The function creates a column for each column name that appears either in the first or in the second data matrix. If a column exists in both data frames, it is row binded as usual. However, if a column is missing in one of the two data frames, the empty cells are `r colorize("replaced by NA", "blue")`.
